Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2023-01-31 paper_comparison/paper_comparison/paper_comparison.Rmd
fix the x range issue

map_df(levels(dat_poly_j$shape), function(x) {

  # Adjust the range of x
  result <- data.frame(log_effect_size = seq(min(dat_poly_j$log_effect_size),
                                             max(dat_poly_j$log_effect_size), 
                                             0.01))

  # Manually fit the model
  conv_mod <- glm(reject ~ log_effect_size, 
                  family = binomial(),
                  data = filter(dat_poly_j, 
                                shape == x, 
                                conv_or_not == "Conventional"))

  # Manually fit the model
  visual_mod <- glm(reject ~ log_effect_size, 
                  family = binomial(),
                  data = filter(dat_poly_j, 
                                shape == x, 
                                conv_or_not == "Visual"))

  # Get the prediction
  result <- result %>%
    mutate(Conventional = predict(conv_mod, 
                                  newdata = result, 
                                  type = "response"),
           Visual = predict(visual_mod, 
                            newdata = result, 
                            type = "response"),
           shape = x) %>%
    pivot_longer(Conventional:Visual, names_to = "conv_or_not")

  result
}) %>%
  ggplot() +
  geom_smooth(aes(log_effect_size, value, col = shape), se = FALSE, span = 0.03) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  scale_color_manual(values = rcartocolor::carto_pal(4, "TealGrn")) +
  theme_light() +
  theme(legend.position = "bottom") +
  facet_wrap(~conv_or_not) +
  labs(col = "Non-linearity shape")
```

@@

map_df(levels(dat_heter_a$a), function(x) {

  # Adjust the range of x
  result <- data.frame(log_effect_size = seq(min(dat_heter_a$log_effect_size),
                                             max(dat_heter_a$log_effect_size), 
                                             0.01))

  # Manually fit the model
  conv_mod <- glm(reject ~ log_effect_size, 
                  family = binomial(),
                  data = filter(dat_heter_a, 
                                a == x, 
                                conv_or_not == "Conventional"))

  # Manually fit the model
  visual_mod <- glm(reject ~ log_effect_size, 
                  family = binomial(),
                  data = filter(dat_heter_a, 
                                a == x, 
                                conv_or_not == "Visual"))

  # Get the prediction
  result <- result %>%
    mutate(Conventional = predict(conv_mod, 
                                  newdata = result, 
                                  type = "response"),
           Visual = predict(visual_mod, 
                            newdata = result, 
                            type = "response"),
           a = x) %>%
    pivot_longer(Conventional:Visual, names_to = "conv_or_not")

  result
}) %>%
  ggplot() +
  geom_smooth(aes(log_effect_size, value, col = a), se = FALSE, span = 0.03) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  scale_color_manual(values = rcartocolor::carto_pal(3, "Earth")) +
  theme_light() +
  theme(legend.position = "bottom") +
  facet_wrap(~conv_or_not) +
  labs(col = "Heteroskedasticity shape")
```

@@

# Extract three datasets, one for poly conventional, one for heter conventional
# and one for visual tests.
list(
poly_conv_sim %>% 
  mutate(log_effect_size = log(effect_size)) %>%
  select(x_dist, log_effect_size, RESET4_p_value) %>%
  rename(RESET_p_value = RESET4_p_value) %>%
  mutate(x_dist = ifelse(x_dist == "even_discrete",
                         "Discrete", 
                         stringr::str_to_title(x_dist))) %>%
  mutate(reject = as.numeric(RESET_p_value < 0.05)) %>%
  mutate(conv_or_not = "Conventional") %>%
  mutate(type = "non-linearity") %>%
  select(type, x_dist, log_effect_size, reject, conv_or_not)
,
heter_conv_sim %>% 
  mutate(log_effect_size = log(effect_size)) %>%
  select(x_dist, log_effect_size, BP_p_value) %>%
  mutate(x_dist = ifelse(x_dist == "even_discrete",
                         "Discrete", 
                         stringr::str_to_title(x_dist))) %>%
  mutate(reject = as.numeric(BP_p_value < 0.05)) %>%
  mutate(conv_or_not = "Conventional") %>%
  mutate(type = "heteroskedasticity") %>%
  select(type, x_dist, log_effect_size, reject, conv_or_not)
,
vi_survey %>%
  filter(!null_lineup) %>%
  filter(!attention_check) %>%
  group_by(unique_lineup_id) %>%
  summarise(across(everything(), first)) %>%
  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
  mutate(reject = as.numeric(p_value <= 0.05)) %>%
  mutate(log_effect_size = log(effect_size)) %>%
  mutate(x_dist = ifelse(x_dist == "even_discrete", 
                         "Discrete", 
                         stringr::str_to_title(x_dist))) %>%
  mutate(conv_or_not = "Visual") %>%
  select(type, x_dist, log_effect_size, reject, conv_or_not)
) %>%

@@

  # Manually control factor ordering
  mutate(type = factor(type, 
                       levels = c("non-linearity", 
                                  "heteroskedasticity"))) %>%
  mutate(x_dist = factor(x_dist, 
                         levels = c("Uniform", 
                                    "Normal", 
                                    "Lognormal", 
                                    "Discrete"))) %>%

@@

  # Bold the uniform smooth line
  geom_smooth(data = ~filter(.x, x_dist == "Uniform"),
              aes(log_effect_size, reject), 
              col = "grey", 
              method = "glm", 
              method.args = list(family = binomial), 
              se = FALSE, 
              size = 3) +

  # Draw all smooth lines
  geom_smooth(aes(log_effect_size, reject, col = x_dist), 
              method = "glm", 
              method.args = list(family = binomial), 
              se = FALSE, 
              size = 1) +

@@

@@

Residual plots rejected by visual tests are rarely not rejected by conventional tests. Throughout the three human subject experiments we conducted, departures of residuals that are of little practical sense to humans will often be alarmed by conventional tests, supporting the common guidance given by experts to utilize graphical representations of data in model diagnostics.

Conventional tests relying on

such as RESET test

1. summarise paper content
2. briefly describe what was learned
3. how does this apply more broadly than simple regression
4. where does this lead? Computer vision residual plot reading

>>>>

dat_poly_j %>%
  mutate(es_min = min(log_effect_size), 
         es_max = max(log_effect_size)) %>%
  nest(data = c(log_effect_size, reject)) %>%
  mutate(mod = map(data, 
                   ~glm(reject ~ log_effect_size, 
                        family = binomial(), 
                        data = .x))) %>%
  mutate(reject = map(mod, function(mod) {
    result <- data.frame(log_effect_size = seq(es_min[1], es_max[1], 0.01))
    result$reject <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-es_min, -es_max, -data, -mod) %>%
  unnest(reject) %>%
  ggplot() +
  geom_smooth(aes(log_effect_size, reject, col = shape), se = FALSE, span = 0.03) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  scale_color_manual(values = rcartocolor::carto_pal(4, "TealGrn")) +
  theme_light() +
  theme(legend.position = "bottom") +
  facet_wrap(~conv_or_not) +
  labs(col = "Non-linearity shape")
```

@@

# Manually fit the model
dat_heter_a %>%
  mutate(es_min = min(log_effect_size), 
         es_max = max(log_effect_size)) %>%
  nest(data = c(log_effect_size, reject)) %>%
  mutate(mod = map(data, 
                   ~glm(reject ~ log_effect_size, 
                        family = binomial(), 
                        data = .x))) %>%
  mutate(reject = map(mod, function(mod) {
    result <- data.frame(log_effect_size = seq(es_min[1], es_max[1], 0.01))
    result$reject <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-es_min, -es_max, -data, -mod) %>%
  unnest(reject) %>%
  ggplot() +
  geom_smooth(aes(log_effect_size, reject, col = a), se = FALSE, span = 0.03) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  scale_color_manual(values = rcartocolor::carto_pal(3, "Earth")) +
  theme_light() +
  theme(legend.position = "bottom") +
  facet_wrap(~conv_or_not) +
  labs(col = "Heteroskedasticity shape")
```

@@

# Extract three datasets, one for poly conventional, one for heter conventional
# and one for visual tests.

dat_x_dist <- list(
  poly_conv_sim %>% 
    mutate(log_effect_size = log(effect_size)) %>%
    select(x_dist, log_effect_size, RESET4_p_value) %>%
    rename(RESET_p_value = RESET4_p_value) %>%
    mutate(x_dist = ifelse(x_dist == "even_discrete",
                           "Discrete", 
                           stringr::str_to_title(x_dist))) %>%
    mutate(reject = as.numeric(RESET_p_value < 0.05)) %>%
    mutate(conv_or_not = "Conventional") %>%
    mutate(type = "non-linearity") %>%
    select(type, x_dist, log_effect_size, reject, conv_or_not),

  heter_conv_sim %>% 
    mutate(log_effect_size = log(effect_size)) %>%
    select(x_dist, log_effect_size, BP_p_value) %>%
    mutate(x_dist = ifelse(x_dist == "even_discrete",
                           "Discrete", 
                           stringr::str_to_title(x_dist))) %>%
    mutate(reject = as.numeric(BP_p_value < 0.05)) %>%
    mutate(conv_or_not = "Conventional") %>%
    mutate(type = "heteroskedasticity") %>%
    select(type, x_dist, log_effect_size, reject, conv_or_not),

  vi_survey %>%
    filter(!null_lineup) %>%
    filter(!attention_check) %>%
    group_by(unique_lineup_id) %>%
    summarise(across(everything(), first)) %>%
    mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
    mutate(reject = as.numeric(p_value <= 0.05)) %>%
    mutate(log_effect_size = log(effect_size)) %>%
    mutate(x_dist = ifelse(x_dist == "even_discrete", 
                           "Discrete", 
                           stringr::str_to_title(x_dist))) %>%
    mutate(conv_or_not = "Visual") %>%
    select(type, x_dist, log_effect_size, reject, conv_or_not)
) %>%

@@

  # Manually control factor ordering
  mutate(type = factor(type, 
                       levels = c("non-linearity", 
                                  "heteroskedasticity"))) %>%
  mutate(x_dist = factor(x_dist, 
                         levels = c("Uniform", 
                                    "Normal", 
                                    "Lognormal", 
                                    "Discrete")))

# Fit the model manually
dat_x_dist %>%
  mutate(es_min = min(log_effect_size), 
         es_max = max(log_effect_size)) %>%
  nest(data = c(log_effect_size, reject)) %>%
  mutate(mod = map(data, 
                   ~glm(reject ~ log_effect_size, 
                        family = binomial(), 
                        data = .x))) %>%
  mutate(reject = map(mod, function(mod) {
    result <- data.frame(log_effect_size = seq(es_min[1], es_max[1], 0.01))
    result$reject <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-es_min, -es_max, -data, -mod) %>%
  unnest(reject) %>%

@@

  # Bold the uniform smooth line
  geom_smooth(data = ~filter(.x, x_dist == "Uniform"),
              aes(log_effect_size, reject), 
              col = "grey", 
              se = FALSE, 
              span = 0.03,
              size = 3) +

  # Draw all smooth lines
  geom_smooth(aes(log_effect_size, reject, col = x_dist), 
              span = 0.03,
              se = FALSE, 
              size = 1) +

@@

1. summarise paper content

@@

2. briefly describe what was learned

Throughout the three human subject experiments we conducted, residual plots rejected by visual tests are rarely not rejected by conventional tests, but departures of residuals that are of little practical sense to humans will often be alarmed by conventional tests, supporting the common guidance given by experts to utilize graphical representations of data in model diagnostics.

We have also found that residual-based conventional tests relying on a user-parameterized test formula, such as the RESET test, will suffer from a decrease in power for detecting certain residual departures intended to be tested but not covered by the test formula. This can not be avoided without prior knowledge of their functional forms, which is usually obtained by plotting the residuals. On the other hand, visual tests do not rely on such information and a collection of departures can be tested simultaneously.

Besides, fitted value distributions

with different

symmetric, evenly

3. how does this apply more broadly than simple regression

4. where does this lead? Computer vision residual plot reading

### 2023-01-31 paper_comparison/paper_comparison/paper_comparison.tex
fix the x range issue

@@

Residual plots rejected by visual tests are rarely not rejected by
conventional tests. Throughout the three human subject experiments we
conducted, departures of residuals that are of little practical sense to
humans will often be alarmed by conventional tests, supporting the
common guidance given by experts to utilize graphical representations of
data in model diagnostics.

Conventional tests relying on

such as RESET test

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  summarise paper content
\item
  briefly describe what was learned
\item
  how does this apply more broadly than simple regression
\item
  where does this lead? Computer vision residual plot reading
\end{enumerate}

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  summarise paper content
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  briefly describe what was learned
\end{enumerate}

Throughout the three human subject experiments we conducted, residual
plots rejected by visual tests are rarely not rejected by conventional
tests, but departures of residuals that are of little practical sense to
humans will often be alarmed by conventional tests, supporting the
common guidance given by experts to utilize graphical representations of
data in model diagnostics.

We have also found that residual-based conventional tests relying on a
user-parameterized test formula, such as the RESET test, will suffer
from a decrease in power for detecting certain residual departures
intended to be tested but not covered by the test formula. This can not
be avoided without prior knowledge of their functional forms, which is
usually obtained by plotting the residuals. On the other hand, visual
tests do not rely on such information and a collection of departures can
be tested simultaneously.

Besides, fitted value distributions

with different

symmetric, evenly

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\item
  how does this apply more broadly than simple regression
\item
  where does this lead? Computer vision residual plot reading
\end{enumerate}
